Biological sequences, such as DNA, RNA and protein, are frequently analyzed in the field of molecular biology and bioinformatics. Multiple sequence alignments (MSAs) help identify similarity between the sequences, and reveal their functional, structural or evolutionary relationships. MSA analysis often requires various visualization techniques, which assists researchers to better understand, evaluate and learn from the MSA results. One of the common techniques is the MSA coloring. There are several coloring tools available \cite{Procter2010aa}, but their color schemes focus more on chemical properties of the sequence residues, rather than the actual structural or evolutionary information of the whole alignment.

In this thesis, we introduce Mavis, a new approach to coloring MSAs and highlighting the quality and structure. Instead of using a pre-defined color scheme based on residue types, we designed a new algorithm to dynamically generate a color for each residue, based on a user-determined similarity scoring between them. This new tool colors an MSA in such a way that a well-aligned region is represented by a solid-color block, while a poor-aligned one by a mosaic with various colors. Thus, the alignment quality and internal structure are clearly displayed.

We implemented the core Mavis algorithm in R, the application programming interface (API) in Perl CGI, and several graphical user interfaces (GUIs) over web browsers and mobile devices like iPhone and iPad. We also followed common software design principles, and the system is extensible for even more add-on features and new user interfaces. The software is ready-to-use for any biologists without requiring deep computer science skills, and will deliver the visualization result of a typical MSA in minutes.